---
title: "Schedule"

output:
  distill::distill_article:
    self_contained: true
    toc: true
    toc_float: true
---

```{r, include=FALSE}
require(here)
require(kableExtra)

knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)

tab <- read.csv(here('resources/schedule.csv'))
```

```{r}

tbl <- function(tab,d){
  
  loc <- which(tab$week==d)
  
  tab2 <- tab[loc,]
  
  
  
  if(!is.na(tab2$slides)){
      
    tab2$slides_      <- paste0(cell_spec(strsplit(tab2$slides,';')[[1]],
                                          'html',
                                          link=strsplit(tab2$slides_link,';')[[1]]),
                                collapse = '<br>')
  } else {
    tab2$slides_ <- ''
  }
  
  
  if(!is.na(tab2$reading)){
     
    tab2$reading_     <- paste0(cell_spec(strsplit(tab2$reading,';')[[1]],
                                            'html',
                                            link=strsplit(tab2$reading_link,';')[[1]]),
                                  collapse = '<br>')
  } else{
    tab2$reading_ <- ''
  }
  
  if(!is.na(tab2$assignments)){
    
    tab2$assignments_ <- paste0(cell_spec(strsplit(tab2$assignments,';')[[1]],
                                            'html',
                                          link=strsplit(tab2$assignments_link,';')[[1]]),
                                  collapse = '<br>')
  } else{
    tab2$assignments_ <- ''
  }
  
  if(!is.na(tab2$date)){
    
    tab2$date_ <- paste0(cell_spec(strsplit(tab2$date,';')[[1]],
                                            'html'),
                                  collapse = '<br>')
  } else{
    tab2$date_ <- ''
  }
  
  
  if(!is.na(tab2$topic)){
    
    tab2$topic_ <- paste0(cell_spec(strsplit(tab2$topic,';')[[1]],
                                            'html'),
                                  collapse = '<br>')
  } else{
    tab2$topic_ <- ''
  }
  
    
  if(!is.na(tab2$lab)){
    
    tab2$lab_ <- paste0(cell_spec(strsplit(tab2$lab,';')[[1]],
                                            'html',
                                          link=strsplit(tab2$lab_link,';')[[1]]),
                                  collapse = '<br>')
  } else{
    tab2$lab_ <- ''
  }
  
  if(!is.na(tab2$r)){
    
    tab2$r_ <- paste0(cell_spec(strsplit(tab2$r,';')[[1]],
                                            'html',
                                          link=strsplit(tab2$r_link,';')[[1]]),
                                  collapse = '<br>')
  } else{
    tab2$r_ <- ''
  }
  
  
  if(!is.na(tab2$data)){
    
    tab2$data_ <- paste0(cell_spec(strsplit(tab2$data,';')[[1]],
                                            'html',
                                          link=strsplit(tab2$data_link,';')[[1]]),
                                  collapse = '<br>')
  } else{
    tab2$data_ <- ''
  }
  
    
  if(!is.na(tab2$week)){
    
    tab2$week_ <- paste0(cell_spec(tab2$week,'html'),
                                  collapse = '<br>')
  } else{
    tab2$week_ <- ''
  }
  
  
  tab2 <- tab2[,c('week_','date_','topic_','slides_','r_','data_','reading_','assignments_','lab_')]
  
  rownames(tab2) <- NULL
  
  tab2 %>%
    kbl(format = 'html',
        escape = FALSE,
        col.names = c('Week','Dates','Topic','Slides','R scripts','Datasets','Readings','Assignments','Lab')) %>%
    kable_material(c("striped", "hover"),full_width = T) %>%
    row_spec(0, background = "#F6FBF4") %>%
    row_spec(0:1, extra_css = "border-bottom: solid;") %>%
    row_spec(0:1, extra_css = "border-top: solid;") %>%
    row_spec(0:1, extra_css = "border-left: solid;") %>%
    row_spec(0:1, extra_css = "border-right: solid;")
  
}
```

### Week 1: Introduction

```{r}

tbl(tab,d=tab$week[1])

```

### Week 2: Describing Continuous Variables

```{r}

tbl(tab,d=tab$week[2])

```


### Week 3: Law of Large Numbers

```{r}

tbl(tab,d=tab$week[3])

```

### Week 4-5: Introduction to Inference

```{r}

tbl(tab,d=tab$week[4])

```

```{r}

tbl(tab,d=tab$week[5])

```

### Week 6: Describing and testing the relationship between two continuous variables

```{r}

tbl(tab,d=tab$week[6])

```

### Week 7: Some history

*No class on Nov 11 (Veterans Day holiday)*

```{r}

tbl(tab,d=tab$week[7])

```

### Week 8-9: Describing and testing the relationship between categorical variables

*No class on Nov 27 (Thanksgiving holiday)*


```{r}

tbl(tab,d=tab$week[8])

```

```{r}

tbl(tab,d=tab$week[9])

```

### Week 10: Wrapping up and critiques of Null Hypothesis Significance Testing

```{r}

tbl(tab,d=tab$week[10])

```

### Week 11: Finals week

Final assignment is due on December 10 at midnight.


